**Best practices for applying machine learning to bacterial 16S rRNA gene sequencing data**

### 2019-09-02 submission/manuscript.Rmd
edit the whole thing

To showcase a rigorous ML pipeline and to shed light on how ML model selection can affect modeling results, we performed an empirical analysis comparing 7 modeling approaches with the same dataset and pipeline. We built three linear models with different forms of regularization: L2-regularized logistic regression and L1 and L2-regularized support vector machines (SVM) with a linear kernel. We also built four non-linear models: SVM with radial basis function kernel, a decision tree, random forest and XGBoost. We compared the predictive performance, interpretability, and computational efficiency. To demonstrate the performance of these modeling approaches and our pipeline, we used data from a previously published study that sought to classifiy individuals as having normal colons or colonic lesions based on the 16S rRNA gene sequences collected from fecal samples [@baxter_dna_2016]. This dataset was selected because it is a relatively large collection of individuals (N=490) connected to a clinically significant disease where there is ample evidence that the disease is driven by variation in the microbiome  [@baxter_microbiota-based_2016; @baxter_dna_2016; @zeller_potential_2014; @knights_human-associated_2011]. With this dataset we developed a framework that implementf a ML pipeline that can be used for any modeling approach, evaluates predictive performance, and demonstrates how to interpret these models. This framework can be easily applied to other host-associated and environmental microbiome datasets.

@@

The foundation of this framework was to clearly define the ML task and to choose the best model for that task [Table S1]. We benchmarked seven ML models and showed that if the ML task is to identify microbial biomarkers of disease, we can create models that are inherently interpretable and easily trained without losing predictive power. In terms of predictive performance, random forest model had the best testing AUROC values compared to the other six models. However, the second best model was L2-regularized logistic regression with a median AUROC difference of only 0.015 compared to random forest. While random forest took 83.2 hours to train, L2-regularized logistic regression trained in 12 minutes. In terms of interpretability, random forest was a complex ML model and could only be explained using post-hoc methods such as permutation importance. On the other hand, L2-regularized logistic regression was easier to interpret by ranking the OTUs based on their feature weights and reviewing the signs of these weights. Comparing many different models showed us that the most complex model is not necessarilty the best model for our ML task.

>>>>

To showcase a rigorous ML pipeline and to shed light on how ML model selection can affect modeling results, we performed an empirical analysis comparing 7 modeling approaches with the same dataset and pipeline. We built three linear models with different forms of regularization: L2-regularized logistic regression and L1 and L2-regularized support vector machines (SVM) with a linear kernel. We also built four non-linear models: SVM with radial basis function kernel, a decision tree, random forest and XGBoost. We compared the predictive performance, interpretability, and computational efficiency. To demonstrate the performance of these modeling approaches and our pipeline, we used data from a previously published study that sought to classifiy individuals as having normal colons or colonic lesions based on the 16S rRNA gene sequences collected from fecal samples [@baxter_dna_2016]. This dataset was selected because it is a relatively large collection of individuals (N=490) connected to a clinically significant disease where there is ample evidence that the disease is driven by variation in the microbiome  [@baxter_microbiota-based_2016; @baxter_dna_2016; @zeller_potential_2014; @knights_human-associated_2011]. With this dataset we developed a framework that implements a ML pipeline that can be used for any modeling approach, evaluates predictive performance, and demonstrates how to interpret these models. This framework can be easily applied to other host-associated and environmental microbiome datasets.

@@

The foundation of this framework was to clearly define the ML task and to choose the best model for that task [Table S1]. We benchmarked seven ML models and showed that if the goal is to identify microbial biomarkers of disease, we can create models that are inherently interpretable and easily trained without losing predictive power. In terms of predictive performance, random forest model had the best testing AUROC values compared to the other six models. However, the second best model was L2-regularized logistic regression with a median AUROC difference of only 0.015 compared to random forest. While random forest took 83.2 hours to train, L2-regularized logistic regression trained in 12 minutes. In terms of interpretability, random forest was a complex ML model and could only be explained using post-hoc methods such as permutation importance. On the other hand, L2-regularized logistic regression was easier to interpret by ranking the OTUs based on their feature weights and reviewing the signs of these weights. Comparing many different models showed us that the most complex model is not necessarilty the best model for our ML task.
